---
title: "Dosing and Body Size"
author: "Ryan Crass"
format: 
    html:
       number-sections: true
       embed-resources: true
       self-contained-math: true
---

# Thesis

The pharmacokinetics of both large and small molecules scale by body weight between (e.g., animal to human dose predictions) and within (e.g., adult to pediatric dose predictions). The general functional form for this scaling relationship is the power model.

$$
Y = x^\gamma
$$

Given most drugs are dosed to target average expoosure (Cavg, AUC), this relationship is most often considered for clearance ($CL_i$) when deriving translational predictions from non-clinical species to humans; however, it can also be applied to volume to target maximal (Cmax) concentration thresholds.

$$
CL_i = \theta_{CL,ref}*(WT_i/WT_{ref})^\alpha
$$

The magnitude of the scaling relationship between the PK parameter and body weight ($\alpha$) and therapeutic index in the target indication determine the optimal dosing paradigm across body size. Assuming that the metric used to describe body weight is the primary covariate of the PK parameter most strongly associated with target exposure in the disease population, one would want to use a metric for $WT_i$ that results in an $\alpha = 1$ in the target population.

To this end some authors have suggested testing a host of body size metrics (e.g., body weight \[WT\], body surface area \[BSA\], adjusted ideal body weight \[AIBW\]) using data from the first-in-human clinical study and selecting the metric that results in $\alpha$ closest to $1$. This is a valid approach; however, it introduces challenges in analysis operations and interpretation of analysis results. First, the analysis will test multiple correlated metrics as covariates with selection based on estimation of a parameter. Furthermore, selection of alternative body size metrics (e.g., \{AIBW\}), may complicate translation to external populations, such as pediatrics, in which these metrics are not validated.

I propose a simplified unified approach to evaluation of approach to assessing a body size metric for dosing to allow linear (e.g., mg / unit body size) calculation of dose amounts. As established previously, the relationship between the primary PK parameter for dosing (e.g., e CL) and body weight can be described using a power function centered on a reference weight (commonly 70 kg), identical to that outlined above:

$$
CL_i  = \theta_{CL,ref}*(WT_i/WT_{ref})^\alpha
$$

This relationship ($\alpha$) can be estimated in a population pharmacokinetic analysis of the first-in-human clinical trial, and in pooled analysis across trials as the clinical database matures. No other body size metrics (BSMs) need to be evaluated as covariates in order to be selected as the BSM for dosing.

This is because at the individiaul level all body size metrics ($BSM_i$) are correlated with body weight ($WT_i$). 

*Core Assumption* The relationship between body weight and body size metrics is adequately described using a power model.

$$
WT_i/WT_{ref} = BSM_i^\beta
$$ 
Thus, if the relationship between body weight ($WT_i$) and the primary PK parameter for dose selection (in this example $CL_i$) is known via the equation above, the relationship between this PK parameter and alternative body size metrics ($BSM_i$) is also known via the relationships described in the equations below.

$$
CL_i = \theta_{CL,ref}*(WT_i/WT_{ref})^\alpha = (BSM_i^\beta)^\alpha = (BSM_i)^\gamma
$$ 
Therefore the PK parameter for dose selection $CL_i$ can be assumed to be related to each $BSM_i$ by the value of $\gamma$, the product of the exponentials for the scaling relationship with body weight $alpha$ and the relationship between body weight and the body size metric $\beta$. Assuming the goal to identify a body size metric to facilitate linear derivation of dose amount (e.g., mg / unit BSM), the target then can be expressed simply a:

$$
\gamma = \alpha * \beta_{bsm} \approx 1 
$$ 

The value of $\beta$ can be estimated from a large population (e.g., NHANES), or a representative sample of the target population, including in some cases the one used to estimate the relationship with body weight ($\alpha$) using population PK analysis. Knowing both parameters for a representative sample of a target population, one can then determine the $BSM$ that is most linearly related to the PK parameter for dosing without covariate analysis of these metrics in population PK analysis. 

This approach can be explored using `NHANES` as a representative dataset of adult human clinical trial participants.

# Set Up

## Packages

```{r pkgs, message=FALSE, warning=FALSE}
rm(list=ls())
library(nhanesA)
library(tidyverse)
library(GGally)
```

## NHANES Data

```{r nhanes, message=FALSE, warning=FALSE}
bmx <- nhanes("BMX")
demo <- nhanes("DEMO")
```


# Body Size Metrics

Alternative body size metrics can be derived from standard anthropometic variables (height, weight, BMI) and demographics (sex). The `NHANES` database will service as the source for representative anthropometric data from adults.

## Process NHANEs Data

First, the NHANEs dataset must be processed to derive an analysis dataset for anthropometric variables. The `SEQN` variable is the subject identifier variable for joining individual domains. 

```{r nhanes-na, message = FALSE, warning = FALSE}
data <- left_join(bmx, demo) %>% 
  filter(!is.na(RIDAGEYR) & !is.na(RIAGENDR) & !is.na(BMXWT) & !is.na(BMXHT)) %>% 
  filter(RIDAGEYR >= 18)
```

## Derive Body Size Metrics

Second, alternative body size metrics ($BSM_i$) must be derived for the analysis.  The equations defining these body size metrics are taken from Pai et al 2012 (https://pubmed.ncbi.nlm.nih.gov/22711238/). 

```{r anthro, message=FALSE, warning=FALSE}
anthro <- data %>% 
  select(SEQN, 
         Age = RIDAGEYR, 
         Sex = RIAGENDR, 
         Weight = BMXWT,
         Height = BMXHT, 
         BMI = BMXBMI) %>% 
  mutate(BSA_Mosteller = signif(sqrt(Weight*Height/3600),3), 
         BSA_DuBois = signif(0.007184 * (Weight^0.425) * (Height^0.725),digits=3),
         IBW = ifelse(Sex == "Male", 
                      signif(50 + 2.3*((Height/2.54)-60),3),
                      signif(45.5 + 2.3*((Height/2.54)-60),3)), 
         AIBW = signif(IBW + 0.4*(Weight-IBW), digits=3),
         FFW = ifelse(Sex == "Male", 
                        signif((0.285 *  Weight) + (12.1 * (Height/100)^2),3),
                        signif((0.287 *  Weight) + (9.74 * (Height/100)^2),3)),
         LBW = ifelse(Sex == "Male", 
                        signif((9270 *  Weight) / (6680 + 216 * BMI),3),
                        signif((9270 *  Weight) / (8780 + 244 * BMI),3))) 
```

Third, the analysis dataset is pivoted longer by gathering all the alternative body size metrics into a single `Body Size Metric` variable and define a separate variable for units.

```{r antrho-long, message=FALSE, warning=FALSE}
anthro_long <- anthro %>% 
  pivot_longer(Weight:LBW, names_to = "Body Size Metric", values_to = "Value") %>% 
  mutate(Unit = case_when(`Body Size Metric` == "BMI" ~ "kg/m2", 
                           grepl("BSA", `Body Size Metric`) ~ "m2",
                           `Body Size Metric` == "Height" ~ "cm",
                           .default = "KG"))
```

Vector of variable names for body size metrics of interest.

```{r params, message=FALSE, warning=FALSE, }
params <- c("Height", "BMI", "BSA_Mosteller", "BSA_DuBois", "IBW", "AIBW", "FFW", "LBW")
```


# Exploatory Data Analysis

Exploratory data analysis is performed to evaluate the distributions of body size metrics, as well as, to evaluate the relationships between these variables and body weight.

## Distributions

First, let's visualize the distributions of the various body size metrics, facetted by unit.

```{r histograms, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
hist <- ggplot(aes(x=Value, color = `Body Size Metric`, fill = `Body Size Metric`), data = anthro_long) +
  geom_density(alpha = 0.7) + 
  facet_wrap(~Unit, scales = "free")+
  theme_bw() 

hist_log <- hist + 
  scale_x_log10(guide = "axis_logticks")

hist 
hist_log
```

## Correlations

Second, let's visualize the correlation between body size metrics. As anticipated, they are all highly correlated. Height is only modestly correlated with weight (|Corr| = 0.45) and ideal body weight (IBW) (|Corr| = 0.4), which is itself derived from sex and height. The remaining body size metrics, including BSA and BMI, are all highly correlated with body weight (|Corr| > ~0.8).

```{r correlation, fig.height=6, fig.width=6, message=FALSE, warning=FALSE}
anthro %>% 
  select("Weight", params) %>% 
  rename("HT" = "Height", 
         "WT" = "Weight",
         "BSA(M)" = BSA_Mosteller, 
         "BSA(D)" = BSA_DuBois) %>% 
  ggpairs()
```


## Log-Log Regression

Third, let's evaluate our core assumption regarding the relationship between body weight and alternative body size metrics. 

*Core Assumption* The relationship between body weight and body size metrics is adequately described using a power model.

First, let's define a function to fit log-log regressions of alternative body size metrics versus to determine $\beta$, the power for the log-linear relationship between the alternative body size metrics and body weight.

```{r weight-metric-fit-summary-function}
sumfit_loglog <- function(exp_var_x,
                          exp_var_y = "Weight", 
                          ref_var_y = 70,
                          data = anthro,
                          method = "normal",
                          ci = 0.95,
                          sigdigits = 3) {
  
  if(!method %in% c("normal", "tdist")) {
    rlang::abort(message = "argument `method` must be 'normal' or 'tdist'")}
  if(!ci %in% c(0.90, 0.95)) {
    rlang::abort(message = "argument `ci` must be 0.90 or 0.95")}

  form <- stats::as.formula(paste(paste0("log(",exp_var_y, "/", ref_var_y, ")"),
                                  "~",
                                  paste0("log(",exp_var_x,")")))
  fit <- stats::lm(form, data)
  
  int <- stats::coef(fit)[[1]]
  est <- stats::coef(fit)[[2]]
  se <- sqrt(diag(stats::vcov(fit)))[[2]]
  z <- stats::qt((1 + ci)/2, (length(fit$residuals)-1))

  tab <- data.frame(
    DV = exp_var_x, 
    IDV = exp_var_y,
    CI = paste0(ci*100, "%"),
    Intercept = signif(int, digits=sigdigits),
    Beta = signif(est, digits = sigdigits),
    StandardError = signif(se, digits=sigdigits),
    LCL_Beta = dplyr::case_when(method == "normal" & ci == 0.95
                                ~ signif(est - 1.96*se, digits=sigdigits),
                                method == "normal" & ci == 0.90 
                                ~ signif(est - 1.64*se, digits=sigdigits),
                                  method == "tdist" ~ signif(est - z*se, digits=sigdigits),
                                .default = NA_real_),
    UCL_Beta = dplyr::case_when(method == "normal" & ci == 0.95 
                                ~ signif(est + 1.96*se, digits=sigdigits),
                                method == "normal" & ci == 0.90 
                                ~ signif(est + 1.64*se, digits=sigdigits),
                                method == "tdist" 
                                ~ signif(est + z*se, digits=sigdigits),
                                .default = NA_real_))

  return(tab)
} 

```

Next, let's fit the regressions across metrics using `map` in the pooled population.

```{r fit}
fit <- map_df(params, sumfit_loglog) 
```

And also stratified by sex.

```{r fit-sex}
fit_m <- data.frame()
fit_f <- data.frame()

for(param in params){
  fit_m <- bind_rows(fit_m, sumfit_loglog(exp_var_x = param, exp_var_y = "Weight", ref_var_y = "70", 
                       data = filter(anthro, Sex == "Male")))
  fit_f <- bind_rows(fit_f, sumfit_loglog(exp_var_x = param, exp_var_y = "Weight", ref_var_y = "70", 
                       data = filter(anthro, Sex == "Female")))
}

fit_sex <- bind_rows(mutate(fit_m, Sex = "Male"), mutate(fit_f, Sex = "Female"))
```

..and define labels and factor ordering.

```{r fit-labels}
fit_labels <- fit %>% 
    mutate(`Beta (95% CI)` = paste0(Beta, " (95% CI: ",LCL_Beta," - ", UCL_Beta,")")) %>% 
    mutate(BSM = factor(DV, 
                      levels = c("Height",
                                "BSA_DuBois" ,
                                "BSA_Mosteller",
                                 "AIBW",
                                 "FFW",
                                 "BMI",
                                 "LBW",
                                 "IBW", 
                                 "Body Weight"))) 
```

```{r fit-sex-labels}
fit_sex_labels <- fit_sex %>% 
      mutate(`Beta (95% CI)` = paste0(Beta, " (95% CI: ",LCL_Beta," - ", UCL_Beta,")")) %>% 
    mutate(BSM = factor(DV, 
                      levels = c("Height",
                                "BSA_DuBois" ,
                                "BSA_Mosteller",
                                 "AIBW",
                                 "FFW",
                                 "BMI",
                                 "LBW",
                                 "IBW", 
                                 "Body Weight"))) 
```

Finally let's define a function to generate log-log concordance plots and annotate our fitted beta (95% CI) values.

```{r plot-concord-weight-fxn}
concordplot <- function(metric, data = anthro,col_fit = "black", shape_point = 21){
  
  mod <- lm(log(data[["Weight"]]/70) ~ log(data[[metric]]))
  
  plot <- ggplot(aes(y = Weight, x = !!sym(metric), col = Sex), data = data) + 
    geom_point(shape = shape_point, alpha_point = 0.5)+
    geom_smooth(aes(y = Weight, x = !!sym(metric)), method = "lm",
                se = FALSE, color = col_fit, inherit.aes = FALSE) +
    geom_smooth(aes(y = Weight, x = !!sym(metric)),method = "loess",
                se = FALSE, color = col_fit, linetype = "dashed", inherit.aes = FALSE) + 
    scale_x_log10(guide = "axis_logticks") +
    scale_y_log10(guide = "axis_logticks") + 
    labs(caption = paste("Beta = ", fit_labels$`Beta (95% CI)`[fit_labels$DV==metric])) +
    theme_bw()
  
  return(plot)
}
```

... and generate the plots to evaluate the core assumption using `map`

```{r plot-concord-weight}
map(params, concordplot)
```

The assumption appears valid! The LOESS and linear regression of the log-transformed data are consistent across the data range, consistent with a largely log-linear relationship between body size metrics and body weight. The possible exceptions to this are fat free weight (FFW) and lean body weight (LWB), which show some deviation from linearity due to the distinct and non-overlapping ranges of values for males and females. Perhaps we should evaluate log-linearity stratified by Sex.

```{r plot-concord-weight-sex-fxn}
concordplot_sex <- function(metric, data = anthro, shape_point = 21) {
  
  mod <- lm(log(data[["Weight"]]/70) ~ log(data[[metric]]))
  
  plot <- ggplot(aes(y = Weight, x = !!sym(metric), col = Sex), data = data) + 
    geom_point(shape = shape_point, alpha_point = 0.5, alpha = 0.3)+
    geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
    geom_smooth(method = "loess", se = FALSE, linetype = "dashed", linewidth =1) + 
    scale_x_log10(guide = "axis_logticks") +
    scale_y_log10(guide = "axis_logticks") + 
    labs(caption = paste("Beta (Male) = ", fit_sex_labels$`Beta (95% CI)`[fit_sex_labels$DV==metric&fit_sex_labels$Sex=="Male"], "\n", paste("Beta (Female) = ", fit_sex_labels$`Beta (95% CI)`[fit_sex_labels$DV==metric&fit_sex_labels$Sex=="Female"]))) +
    theme_bw()
  
  return(plot)
}
```


```{r plot-concord-weight-sex}
map(params, concordplot_sex)
```

# Integration of Alpha, Beta, and Gamma

Plot the relationship between values of $\alpha$ (`y`) to be fitted from clinical data for molecules and the fitted values of $\beta$ \* $\alpha$ (`x * y`).

```{r plot-concord-exp, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
ggplot(aes(x = Alpha, y = Gamma, col = BSM), data = data.frame(Alpha = c(0, 2), 
                                                               Gamma = c(0, 2),
                                                               BSM = c(NA, NA))) + 
  geom_abline(aes(intercept = 0, slope = Beta, col = BSM), data = fit_labels) + 
  geom_hline(yintercept = 1) + 
  geom_hline(yintercept = c(0.8, 1.2), linetype = "dashed") + 
  scale_x_continuous(limits = c(0, 2), breaks = c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 2)) + 
  scale_y_continuous(limits = c(0, 2), breaks = c(0, 0.5, 1, 1.5, 2)) + 
  labs(y = expression(paste("Dosing Body Size Metric Exponent (", gamma, ")")), 
       x = expression(paste("Body Weight Exponent (", alpha, ")")), 
       col = expression(paste("Body Size Metric (slope = ", beta, ")")))+
  theme_bw() + 
  theme(panel.grid.major.y = element_blank(), 
        panel.grid.minor.y = element_blank())
```
